Need Fulfillment During Intergroup Contact: Three Experience Sampling Studies

### 2023-05-26 Methods-and-Results.Rmd
power sim re-added and robustness analyses extended (revision)

\subsubsection{Contact Hypothesis}
Using a multilevel regression, we find that having an outgroup contact is indeed associated with significantly more positive outgroup attitudes (`r MlCoeffLatex(lmeMdl = mdlWorker$lmeInterceptAttType, lmerCI = mdlWorker$lmerInterceptAttTypeCI, varName = "OutgroupInteractionC")`), even after controlling for having an interaction with a non-Dutch (which did not relate to outgroup attitudes independently). Additionally, while multi-level regressions are generally robust against unequal cell sizes, we correct for inequalities by using centered predictors and reintroducing the means as level two predictors (\citealp{Yaremych2021}; for full results see \tblref{tab:intergroupGeneralTblLong}, \fgrref{fig:ContactHypothesis}, and Online Supplementary Material A)\footnote{Interestingly, adding random slopes to this model did not explain additional variance. This is unusual and might indicate that the effect is very consistent across participants. However, the small number of participants, or other measurement issues provide an alternative explanation, which is why we offer a combined data set analyses following the individual studies.}. 
Thus, in our first data, we find initial evidence that outgroup contacts show a positive effect on outgroup attitudes within real-life data.

@@

\section{Robustness, Stability, and Embeddedness across Studies}

Beyond the individual results of the three studies, we conducted three additional analyses to test the broader cross-study claims and contextualize our results. In particular, we aimed to gain a broader picture of the individual results by considering them jointly. We also leveraged the data from all participants to test between-participant contact effects and explore the content of the participants' situational contact needs.

\subsection{Stability}
We first assessed the stability of our main analyses across the three studies. Plotting the effect sizes of each parameter of interest in a forest plot, as well as the average meta-analytic effect, shows that for the basic contact hypothesis test outgroup contact had a strong and consistent effect on outgroup attitudes. Interactions with non-outgroup members consistently had no meaningful effect on outgroup attitudes (\fgrref{fig:ContactHypothesis}). While this would be expected from the general intergroup contact literature, this is not a trivial finding. Being among the first to assess the contact hypothesis using real-life intensive longitudinal data, we extend cross-sectional findings to individual-level assessments. When looking at the fulfillment of core needs during intergroup contacts, we find that the motivational mechanism is consistently a meaningful predictor of outgroup attitudes and interaction quality perceptions (see \fgrref[ A and B]{fig:AllportNeedFulfillment}). We also see that the effect of core need fulfillment on outgroup attitudes is strongly reduced when modeled together with interaction quality perceptions, supporting our assertion that interaction quality offers a psychological mechanism for the direct effect (see \fgrref[-C]{fig:AllportNeedFulfillment}).

\subsection{General contact hypothesis}
```{r join data sets for analysis, include=FALSE}
# Join aggregated data
avS1 <- dtWorkerSupp$workerAvFreQual %>%
  select(
    -ExternalReference
  ) %>%
  mutate(
    PID = c(1:nrow(.)),
    study = "S1"
  ) %>%
  select(PID, everything())
avS2 <- dtStudentSupp$studentContactFreq %>%
  mutate(
    study = "S2"
  )
avS3 <- dtMedicalSupp$medicalContactFreq %>%
  mutate(
    study = "S3"
  )

@@

During the main analyses, we have thus far shown that participants held more positive outgroup attitudes following intergroup contacts and that perceived interaction quality was associated with more positive outgroup attitudes following an intergroup contact. We have, however, not brought the two elements of the contact hypothesis together in a single analysis. To jointly test the effects of contact frequency and average interaction quality, we ran a linear regression model where average outgroup attitudes were predicted by the number of interactions and the average interaction quality ratings of all participants across the three studies. We did so while controlling for the possible effects of study-specific differences. To include the study-membership as a control variable, we used the student sample (Study 2) as the reference group because it was both the largest and the most homogeneous study. Looking at the overall model, we found that the model predicted `r format(round(fit$r.squared*100, 2), nsmall = 2)`\% of the variance in average outgroup attitudes (`r lmAllFstatLatex`). Looking at the individual effects, we found that only the number of outgroup interactions has a clear association with average outgroup attitudes (`r LmCoeffLatex(lmMdl = lmAllAttFreqQualX, varName = "SumContactNL_c", ci = TRUE)`). The average interaction quality perceptions had a much smaller effect (`r LmCoeffLatex(lmMdl = lmAllAttFreqQualX, varName = "AvQualityOut_c", ci = TRUE)`), and importantly we found no interaction effect at all. In short, the effect of interaction frequency did not depend on the average interaction quality.

\subsection{Contact key need content}
Finally, we used the qualitative data from the participants’ self-identified core needs to contextualize the results of our main analysis. However, because our participants jointly reported on thousands of intergroup contacts, it would not have been feasible to analyze these qualitative responses in a traditional qualitative content analysis. We instead relied on recent machine learning advances within the natural language processing domain. For our analysis, we used the BERT language model. BERT (Bidirectional Encoder Representations from Transformers) was developed by Google in 2018 and today forms a key element of many natural language processing workflows. In its essence, BERT is a framework that allows users to codify every word in relation to every other word within a large set of documents. We extracted 47 topics from the 2,983 interaction goal free-text entries --- a relatively large number of topics. The higher number of topics allowed us to retain more of the smaller topics and leaves a relatively low number of 308 free-text entries unclassified (10.33%). A full write-up of the topic modeling process is available in Online Supplemental Material B.

In terms of the content of the topics, we find that a number of topics are primarily task-oriented, where participants hope to increase their study, research, presentation, or work performance. Opposing the task- and work-oriented needs, are a wide variety of leisure-related needs wishes, like relaxation and entertainment. Additionally, some clusters were primarily relationship-oriented, so that participants sought contact with outgroup members for intimate and casual social contact in itself. Similarly, socializing and celebrations were also explicit social needs (incl., parties). This also included a subtopic of spiritual, religious, and otherwise transcendental needs (incl. meditation, prayer, religious services). Among the leisure-oriented topics was also a set of contact goals that were specifically migration-specific (e.g., wish to learn about culture, politics, and language) or were concerned with informational needs more generally (e.g., seeking answers, bureaucratic information). A similar set of topics was specifically geared towards a wish to experience cultural products (e.g., music, theater, food) or had travel-related goals in their interactions with the majority group members.

One interesting observation was the importance of contact goals specific to contact through the medical and public health system. This goal type was partly specific to the medical professionals (e.g., working with patients, treatment), but also more broadly highlighted interactions of newcomers with the outgroup majority as patients themselves (e.g., therapeutic goals). Also health, fitness, and personal improvement goals (e.g., sports and music) were common goals that participants shared during the interactions with the majority group members. Another topic less commonly found within the literature were more practically oriented needs, where participants sought to share food and cook together or had organizational needs (e.g., housing, cleaning, and co-living). In sum, almost all extracted topics fall into broader or narrower need concepts that are commonly discussed within the psychological needs literature \citep[e.g.,][]{orehek2018a} and offer insight into a core aspect of the migration experience that has remained broadly under-explored \citep[][]{Kreienkamp2022d}\footnote{It should be noted that these topic models are not without limitations, especially because they depend on a small set of hyperparameters that determine the characteristics of the embedding, dimension reduction, clustering, and core term extraction. The authors have also inspected a major subset of the free-text responses manually and are under the impression that the topics described here accurately represent the broader content.}.

>>>>

\subsubsection{Contact Hypothesis}
Using a multilevel regression, we find that having an outgroup contact is indeed associated with significantly more positive outgroup attitudes (`r MlCoeffLatex(lmeMdl = mdlWorker$lmeInterceptAttType, lmerCI = mdlWorker$lmerInterceptAttTypeCI, varName = "OutgroupInteractionC")`), even after controlling for having an interaction with a non-Dutch (which did not relate to outgroup attitudes independently). Additionally, while multi-level regressions are generally robust against unequal cell sizes, we correct for inequalities by using centered predictors and reintroducing the means as level two predictors (\citealp{yaremych2021a}; for full results see \tblref{tab:intergroupGeneralTblLong}, \fgrref{fig:ContactHypothesis}, and Online Supplementary Material A)\footnote{Interestingly, adding random slopes to this model did not explain additional variance. This is unusual and might indicate that the effect is very consistent across participants. However, the small number of participants, or other measurement issues provide an alternative explanation, which is why we offer a combined data set analyses following the individual studies.}. 
Thus, in our first data, we find initial evidence that outgroup contacts show a positive effect on outgroup attitudes within real-life data.

@@

\section{Stability, Robustness, and Embeddedness across Studies}

Beyond the individual results of the three studies, we conducted a number additional analyses to test the broader cross-study claims, account for alternative models, and contextualize our results. Jointly, these stability, robustness, and embeddedness analyses seek to strengthen our confidence in the results.

\subsection{Stability}
We ran two analyses that tested the stability of our results. We first assess the consistency of the results reported in the three studies and use a meta-analytic approach to gauge the general effect sizes. The second stability analysis we conduct seeks to asses the extend to which the within person contemporaneous effects extend to an aggregated version that mirrors the many cross-sectional recall studies.

\subsubsection{Consistency}
We first assessed the stability of our main analyses across the three studies. Plotting the effect sizes of each parameter of interest in a forest plot, as well as the average meta-analytic effect, shows that for the basic contact hypothesis test outgroup contact had a strong and consistent effect on outgroup attitudes. Interactions with non-outgroup members consistently had no meaningful effect on outgroup attitudes (\fgrref{fig:ContactHypothesis}). While this would be expected from the general intergroup contact literature, this is not a trivial finding. Being among the first to assess the contact hypothesis using real-life intensive longitudinal data, we extend cross-sectional findings to individual-level assessments. When looking at the fulfillment of core needs during intergroup contacts, we find that the motivational mechanism is consistently a meaningful predictor of  interaction quality perceptions and outgroup attitudes (see \fgrref[ A and B]{fig:AllportNeedFulfillment}). We also see that the effect of core need fulfillment on outgroup attitudes is strongly reduced when modeled together with interaction quality perceptions, supporting our assertion that interaction quality offers a psychological mechanism for the direct effect (see \fgrref[-C]{fig:AllportNeedFulfillment}).

<!-- \subsection{General contact hypothesis} -->
\subsubsection{Aggregated Analysis}
```{r join data sets for analysis, include=FALSE}
# Join aggregated data
avS1 <- dtWorkerSupp$workerAvFreQual %>%
  select(
    -ExternalReference
  ) %>%
  mutate(
    PID = c(1:nrow(.)),
    study = "S1"
  ) %>%
  select(PID, everything())
avS2 <- dtStudentSupp$studentContactFreq %>%
  mutate(
    study = "S2"
  )
avS3 <- dtMedicalSupp$medicalContactFreq %>%
  mutate(
    study = "S3"
  )

@@

A second stability test is checking whether the within person effects translate to a broader aggregated between subjects effect, which would mirror common cross-sectional practices. During the main analyses, we have thus far shown that participants held more positive outgroup attitudes following intergroup contacts and that perceived interaction quality was associated with more positive outgroup attitudes following an intergroup contact. We have, however, not brought the two elements of the contact hypothesis together in a single analysis. To jointly test the effects of contact frequency and average interaction quality, we ran a linear regression model where average outgroup attitudes were predicted by the number of interactions and the average interaction quality ratings of all participants across the three studies. We did so while controlling for the possible effects of study-specific differences. To include the study-membership as a control variable, we used the student sample (Study 2) as the reference group because it was both the largest and the most homogeneous study. Looking at the overall model, we found that the model predicted `r format(round(fit$r.squared*100, 2), nsmall = 2)`\% of the variance in average outgroup attitudes (`r lmAllFstatLatex`). Looking at the individual effects, we found that only the number of outgroup interactions has a clear association with average outgroup attitudes (`r LmCoeffLatex(lmMdl = lmAllAttFreqQualX, varName = "SumContactNL_c", ci = TRUE)`). The average interaction quality perceptions had a much smaller effect (`r LmCoeffLatex(lmMdl = lmAllAttFreqQualX, varName = "AvQualityOut_c", ci = TRUE)`), and importantly we found no interaction effect at all. In short, the effect of interaction frequency did not depend on the average interaction quality.

\subsection{Robustness}
To check for spurious relationships, we test three additional models. These three models assess whether the effect is indeed specific to intergroup contact, whether the need fulillment effect is a affected by whether the contact was planned or accidental, and whether the motivational mechanism also holds direct well-being benefits for the minority members. We describe the full methods and results for the robustness analyses in \appref{app:AppendixRobustness}. We use a two-staged analysis approach for the robustness analyses. We first test the model `globally' --- across the three studies --- in a three level multilevel regression (i.e., measurements nested within participants, and participants nested within studies). Only in a second step do check for study-specific idiosyncrasies.

\subsubsection{Contact specific}
To ensure that the core need fulfillment is outgroup contact specific, we return to the full sample of intensive longitudinal measurements and test whether there is an interaction effect of outgroup contact (vs. no outgroup contact) and core need fulfillment. We expected that the effect of core need fulfillment on outgroup attitudes is specific to outgroup interactions and not merely due to a more need-fulfilled life in general. We find that \hl{...}.

\subsubsection{Interaction intend}
To test whether the need mechanism was affected by whether the interaction was accidental or planned we ran an exploratory moderation analysis using the participants' ratings of how much they perceived the interaction as 'accidental'. It should be noted at this point, that in order to keep the ESM surveys short, we asked our participants to focus on most the important interaction (i.e., "\textit{The following questions will be about the interaction \underline{you consider most significant}.}"''"; emphasis as in original). This was also reflected in a relatively low average and consistent right skew of the 'accidental' item. Nonetheless, there remained a substantial variance in the item and we continued with the moderation analysis. We find that \hl{...}.

\subsubsection{Well-being outcome}
Given the well-established criticism that more positive outgroup attitudes might not always be beneficial for minority group members \citep[e.g.,][]{reimer2023}, we conducted an additional exploratory analysis assessing the effect of need fulfilling outgroup interactions on reported well-being. Experienced well-being is a common indicator of health and an important life quality measurement in itself, especially for migrants or other minority groups \citep[e.g.,][]{bhugra2011}. We find that \hl{...}.
The analysis, thus, adds weight to the importance of considering the need fulfillment experiences when it comes to outgroup interactions.

\subsection{Embeddedness}
To embed our results further, we also considered the content and types of needs that participants reported during the study. For these embeddedness analyses, we first assessed the self-reported motives of the participants and then also control the commonly considered self-determination theory needs.

\subsubsection{Contact need content}
We used the qualitative data from the participants’ self-identified core needs to contextualize the results of our main analysis. However, because our participants jointly reported on thousands of intergroup contacts, it would not have been feasible to analyze these qualitative responses in a traditional qualitative content analysis. We instead relied on recent machine learning advances within the natural language processing domain. For our analysis, we used the BERT language model. BERT (Bidirectional Encoder Representations from Transformers) was developed by Google in 2018 and today forms a key element of many natural language processing workflows. In its essence, BERT is a framework that allows users to codify every word in relation to every other word within a large set of documents. We extracted 47 topics from the 2,983 interaction goal free-text entries (after duplicate removal) --- a relatively large number of topics. The higher number of topics allowed us to retain more of the smaller topics and leaves a relatively low number of 308 free-text entries unclassified (10.33%). A full write-up of the topic modeling process is available in Online Supplemental Material C.

In terms of the content of the topics, we find that a number of topics are primarily task-oriented, where participants hope to increase their study, research, presentation, or work performance. Opposing the task- and work-oriented needs, are a wide variety of leisure-related needs wishes, like relaxation and entertainment. Additionally, some clusters were primarily relationship-oriented, so that participants sought contact with outgroup members for intimate and casual social contact in itself. Similarly, socializing and celebrations were also explicit social needs (incl., parties). This also included a subtopic of spiritual, religious, and otherwise transcendental needs (incl. meditation, prayer, religious services). Among the leisure-oriented topics was also a set of contact goals that were specifically migration-specific (e.g., wish to learn about culture, politics, and language) or were concerned with informational needs more generally (e.g., seeking answers, bureaucratic information). A similar set of topics was specifically geared towards a wish to experience cultural products (e.g., music, theater, food) or had travel-related goals in their interactions with the majority group members. In sum, almost all extracted topics fall into broader or narrower need concepts that are commonly discussed within the need fulfillment literature \citep[e.g.,][]{orehek2018a} and offer insight into a core aspect of the migration experience that has remained broadly under-explored \citep[][]{Kreienkamp2022d}\footnote{It should be noted that these topic models are not without limitations, especially because they depend on a small set of hyperparameters that determine the characteristics of the embedding, dimension reduction, clustering, and core term extraction. The authors have also inspected a major subset of the free-text responses manually and are under the impression that the topics described here accurately represent the broader content.}.

<!-- One interesting observation was the importance of contact goals specific to contact through the medical and public health system. This goal type was partly specific to the medical professionals (e.g., working with patients, treatment), but also more broadly highlighted interactions of newcomers with the outgroup majority as patients themselves (e.g., therapeutic goals). Also health, fitness, and personal improvement goals (e.g., sports and music) were common goals that participants shared during the interactions with the majority group members. Another topic less commonly found within the literature were more practically oriented needs, where participants sought to share food and cook together or had organizational needs (e.g., housing, cleaning, and co-living).  -->

\subsubsection{Need types}
To ensure that our results are not impacted by differences in the reported goals and motives, we additionally coded the topics we extracted during the topic modeling on two dimensions of how much they reflect a practical and a psychological goal-directedness. We chose practical and psychological needs specifically as our dimensions to account for differences in the types of needs that participants commonly reported. With practical motives we refer to specific, tangible goals or tasks that participants aimed to accomplish during the interaction. These instrumental goals are usually observable, concrete, and often centered on external outcomes, such as acquiring resources, completing tasks, and addressing immediate challenges \citep[][]{}. With psychological motives we refer to underlying motives or desires that are more abstract and relate to personal fulfillment and well-being. In contrast to practical needs, psychological needs delve into the subjective and internal aspects of human experiences. These needs pertain to emotions, social connections, and cognitive processes, reflecting individuals' quest for personal growth, well-being, and thriving in social relationships \citep[][]{}. Note that with this approach any particular motive can include a practical and/or a psychological goal-directedness but can also be classified as not having any goal at all.

For each of the topics, we had identified in the topic modeling analysis, we extracted the top 10 most indicative words as well as the three most representative free text responses. Using a detailed coding protocol for both the practical as well as the psychological motives, two independent coders rated each of the 47 topics. This allowed us to add the two dimensions back to each of the reported interaction motives and assess the distributions and effects of the goal-directed character. We found that \hl{...}.

\subsubsection{Specific psychological needs}
Finally, to ensure that a much simpler model of three fundamental psychological needs might not account for the same effect, we compared our situation core need fulfillment measurement to the three self-determination theory needs (autonomy, relatedness, competence).
